---
title: "Assessing the sensitivity of results you see in print"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sensitivity_you_see_in_print}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 4.5,
  fig.width = 4.5,
  fig.align = 'center',
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this section we discuss how to use the individual functions of `sensemakr` to peform sensitivity analyses of results you see in print, but do not have access to the microdata. Here we take a more practical approach, and spend less time discussing the intepretation or theory. Therefore, if this is you first time using these tools, it may be helpful to read the vignette "An introduction to sensitivity analysis using `sensemakr`" firt, as well as the JRSS-B paper ( [Cinelli and Hazlett (2020)](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/rssb.12348). )

## Revisiting Darfur

Suppose that, instead of having access to the raw data of the conflicts in Darfur (see `?darfur`), we onlt had instead, access to the usual regression table as presented in a published paper (here we show just the coefficients of `directlyharmed` and `female`, to save space):


```{r, echo=FALSE, message = FALSE}
# loads package
library(sensemakr)

# loads data
data("darfur")

darfur.model <- lm(peacefactor ~ directlyharmed  + village +  female +
                     age + farmer_dar + herder_dar + pastvoted + hhsize_darfur, 
                   data = darfur)
```


```{r, echo=FALSE,  comment = ""}
stargazer::stargazer(darfur.model, 
                     keep = c("directlyharmed","female"),
                     type = "text")
```

Further suppose that the paper claims that there is a large, positive causal effect of `directlyharmed` on `peacefactor`,  but does not present any sensitivity analysis to examine the robustness of such claim against the threat of unobserved confounding. Can we assess how robust these findings are by using just the summary statistics of the regression table? Fortunately, the answer is "yes." Here we show how to replicate all results o the previous vignette "An introduction to sensitivity analysis using `sensemakr`" using the "lower level" functions of `sensemakr`.

## Sensitivity statistics for routine reporting


It is possible to compute the partial $R^2$ and robustness value of `directlyharmed` by simply providing the t-statistics and the degrees of freedom of the regression. The partial $R^2$ is computed using the `partial_r2()` function.


```{r}
partial_r2(t_statistic = 0.097/0.023, dof = 783)
```
The partial $R^2$ here is 2.2%, which means that unobserved confounders that explain 100% of the residual variation of the outcome (`peacefactor`), would still need to explain 2.2% of the the treatment (`directlyharmed`) to bring the point estimate to zero.

The robustness value is computed with the `robustness_value()` function. By default, the function computes the minimum amount of confounding to bring the point estimate (`alpha=1`, ie, significance level of 1) to zero (`q=1`, that is, a reduction of 100% of the current estimate).

```{r}
robustness_value(t_statistic = 0.097/0.023, dof = 783)
```
This means that confounders that explain 13.9% of the residual variance of both the treatment and the outcome are sufficiently strong to explain away the observed estimate of 0.097. Confounders with both associations lower than 13.9% are not sufficiently strong to do so.

These defaults can be changed. For example, the code below computes the minimum strength of confounding to bring the lower end of a 95% confidnce interval (`alpha =0.05`) to half of the current estimate  `q=1/2`. We see that a confounder that explains just 0.05% of both the treatment and the outcome is sufficiently strong to make the confidence interval include the value of 0.0485 (half of the current estimate of 0.97).

```{r}
robustness_value(t_statistic = 0.097/0.023, dof = 783, q = 1/2, alpha = 0.05)
```

It is also possible to compute all these sensitivity statistics with the convenience function `sensitivity_stats()`

```{r}
sensitivity_stats(estimate = 0.097, se = 0.023, dof = 783)
```

## Adjusted estimates and standard errors




```{r}
ovb_contour_plot(estimate = 0.097, se = 0.023, dof = 783)
```

