---
title: "The sensemakr vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The sensemakr vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# setup

```{r}

knitr::opts_chunk$set(echo = TRUE,
                      fig.width=6,
                      fig.height=6)
```

## Install ‘sensemakr’

```{r}

# devtools::install_github("chadhazlett/sensemakr")

# setwd("~/projects/sensemakr_fin/sensemakr/")

# devtools::build_vignettes() 

```

## Darfur Data Example


```{r}
library(sensemakr)
data("darfur")
```

## The 3 Steps

1) Fit a linear model via `lm()`
2) Compute sensitivity quantitites via `sensemakr()`
3) Plot quantities via `plot()`


# Continuous Benchmarks

## Fit

```{r}
model_cntns = lm(data = darfur,
                 peacefactor ~ directlyharmed +
                   age + hhsize_darfur)

```

## Compute

```{r}
sense_cntns = sensemakr(model=model_cntns,
                        treatment="directlyharmed")

str(sense_cntns$benchmarks,max.level = 1)

class(sense_cntns)

```

## Plot
```{r}
plot(sense_cntns,lim=0.02)
```

### Show specific benchmark points from an explicit list



```{r}
plot(sense_cntns,showvars=list('age','hhsize_darfur'),lim=0.02)
plot(sense_cntns,showvars=list('age'),lim=0.02)
```

# Factor Benchmarks

## Fit

```{r}

model_fctr  = lm(data = darfur,
                 peacefactor ~ directlyharmed + 
                   age + hhsize_darfur +
                   female + village)

```

## Compute


```{r}


sense_fctr = sensemakr(model=model_fctr,
                       treatment="directlyharmed")


```

notice there is a single observation in the `$benchmark_group` item of the `sensemakr` object

```{r}
str(sense_fctr$benchmarks,max.level = 1)

sense_fctr$benchmarks$benchmark_group

row.names(sense_fctr$benchmarks$benchmark_group)

```

Village enters the linear model as a R-side ‘factor’ variable with 480 levels. These 480 levels are dummy variable columns of the design matrix. Numerically, each level is treated as a set of binary covariates (fixed effects), and so the sensemakr() call treats each of the 480 levels of the village factor as a stand alone dummy variable which results in 480 benchmark points.

These 'low-level' quantities are stored in the `$benchmarks$benchmark_eachvar` item of the `sensemakr` object

```{r}

str(sense_fctr$benchmarks$benchmark_eachvar)


```

Although we have 480 dummy variable sensitivity estimates for the village factor levels, we would like to understand the sensitivity when benchamrked to the entire single factor variable as a group

notice that we have also grouped all 480 levels into the single `village` term stored in `str(sense_fctr$benchmarks$benchmark_group)`

```{r}

str(sense_fctr$benchmarks$benchmark_group)

```
We have also masked out the 480 levels from `str(sense_fctr$benchmarks$benchmark_masked)`

```{r}
str(sense_fctr$benchmarks$benchmark_masked)

```

## Plot

Show all low level (design matrix) benchmark points


```{r}

plot(sense_fctr,showvars='all',lim=0.02)


```

Above, we immediately see the reason why using many dummy variables (for a single factor) can pose a challenge for interpretation, visually and contextually.

Contextually, binary contrast columns (for each level) should be viewed as a single grouped ‘village’ factor. Visually, we only want to plot the single `village` factor and mask out its associated 480 individual levels.

Therefore, we have designated `showvars='masked'` as the default option in `plot(...,showvars='masked')`

```{r}

# plot(sense_fctr,showvars='masked',lim=0.02)
# same as

plot(sense_fctr,lim=0.02)

```

### Show specific benchmark points from an explicit list

Show the three benchmarks:

* the `village` factor group
* the `villageMngao` level of the village factor
* the continuous covariate `age`



```{r}

par(mfrow=c(2,2))

plot(sense_fctr,showvars=list('village','villageMngao','age'),lim=0.5)


plot(sense_fctr,showvars=list('village','villageMngao','age'),lim=0.02)


```


# User Specified Groups to Benchmark

## Fit

```{r}
model_cust_grp  = lm(data = darfur,
                     peacefactor ~ directlyharmed + 
                       age + hhsize_darfur +
                       female + village
                     )

```

notice for this example, we are using the same resulting model fit that was used in the previous example

```{r}
identical(model_cust_grp,model_fctr)

```

## Compute

Although the model fit step was similar, if the user wishes to compute sensitvitiy quantities for a custom grouping, the grouping must be specified during the `sensemakr()` call using the optional `group_list` argument. As in `sensemakr(...,group_list = list(c('village','female'),'age'))`.

The example `group_list = list(c('village','female'),'age')` has two groups

1) `village` with `female`
2) `age` by itself

The `group_list` argument expects a list of character vectors. The terms in a character vector comprise a single group. 


```{r}

sense_cust_grp = sensemakr(model=model_cust_grp,
                           treatment="directlyharmed",
                           group_list = list(c('village','female'),'age')
                           )

str(sense_cust_grp$benchmarks,max.level = 1)
```

The sensitivity quantities are stored in the `$benchmarks$benchmark_group` item of the `sensemakr` object

```{r}
names(sense_cust_grp$benchmarks)

head(sense_cust_grp$benchmarks$benchmark_masked)
head(sense_cust_grp$benchmarks$benchmark_group)
```




Terms, eligible for grouping, must be the same terms present in the initial outcome regression model stored in the `lm` object.

```{r}
class(model_cust_grp)

colnames(attr(terms(formula(model_cust_grp)),'factor'))

```




## Plot

### Plot with showvars=‘all’

```{r}
plot(sense_cust_grp,showvars='all',lim=0.5)
```


### Plot with showvars=‘masked’

```{r}
plot(sense_cust_grp)
```

From the plot (with default showvars='masked'), notice that:

* ‘village,female’ is plotted 
* ‘female’ is not plotted, 
* ‘village’ (group) is plotted, 
* the 480 factor levels of ‘village’ are not plotted
* 'age' is plotted as cyan

Depending on the showvars option, the plot methods `plot(...,showvars='masked')` (default) or `plot(...,showvars='all')` will only display two sets of points, 

1) `benchmark_group` (both showvars options) 
and
2) either of `benchmark_masked` or `benchmark_eachvar`

The terms in `benchmark_masked` are plotted if `plot(...,showvars='masked')`.

The terms in `benchmark_eachvar` are plotted if `plot(...,showvars='all')`

If a term is in `benchmark_eachvar` that is also part of a term in `benchmark_group`, then it does **not** get copied into `benchmark_masked`.


The `female` term is stored in `benchmark_eachvar` and is also part of the `c("village","female")` group. Therefore `female` is **not** stored in `benchmark_masked` and is not displayed in the default `plot(...,showvars='masked')`.

The factor levels of `village` are not stored in `benchmark_masked` since they are in `benchmark_eachvar` but also part of the group `village`.

The `village` term itself is a standalone group comprised of 480 levels hence present in `benchmarks_group` . In the examples before, we showed that the internals of `sensemakr()` enforced factor levels to be treated as a group.

The single term `age` is part of the second user specified group 
`group_list = list(c('village','female'),'age')` .


More generally, all sensitivity quantities, related to the columns of the linear model's model matrix, are stored in `benchmark_eachvar`. All terms in `benchmark_eachvar` get plotted if `plot(...,showvars='all')`.

If a term is in `benchmark_eachvar` and is also part of a term in `benchmark_group`, then that redundant term is **not** stored in `benchmark_masked`. That is, `benchmark_masked` is the set complement of `benchmark_eachvar` anti-joined against `benchmark_group`. These redundant terms do not get plotted if `plot(...,showvars='masked')`.

If a term is in `benchmark_group`, it is not eligible to be copied into `benchmark_masked`.


If the user wishes to mask specific terms in `benchmark_group` they must do it explicily themselves, say if the user wants to plot `village,female` (a user specified group) but not `village` (a sensemakr-enforced group). 


### Plot with showvars=list(‘village,female’)

When choosing explicit groups to plot with the `showvars` argument of `plot()`, you must supply a `list()` whose entries are single character vectors. A single string must concatenate all the terms belonging to a single group (seperated by a spaceless comma).

For example, the plot command `plot(...,showvars=list(‘village,female’))` corresponds with 

the compute command `sensemakr(...,group_list = list(c('village','female')))`


```{r}

plot(sense_cust_grp,showvars=list('village,female'),lim=0.5)

```

notice how `village,female` is plotted but **not** `village`.



# Other Helpful Plots

## Other Contours

with default showvars=‘masked’

```{r}

par(mfrow=c(2,2))

plot(sense_cust_grp, lim=.2)
plot(sense_cust_grp, contour = "t-value")
plot(sense_cust_grp, contour = "lower bound")
plot(sense_cust_grp, contour = "upper bound")

```

with showvars=‘all’

```{r}

par(mfrow=c(2,2))

plot(sense_cust_grp, contour = "t-value",showvars='all')
plot(sense_cust_grp, contour = "t-value",showvars='all',lim=0.02)
plot(sense_cust_grp, contour = "lower bound",showvars='all',lim=0.02)
plot(sense_cust_grp, contour = "upper bound",showvars='all',lim=0.02)

```


with showvars explicit


```{r}

par(mfrow=c(2,2))

plot(sense_cust_grp, contour = "t-value",
     showvars=list('village','villageMngao','age'))

plot(sense_cust_grp, contour = "lower bound",
     showvars=list('village','villageMngao','age'))

plot(sense_cust_grp, contour = "upper bound",
     showvars=list('village','villageMngao','age'))
```


### Worst Case Plot

```{r}

par(mfrow=c(2,2))

plot(sense_cust_grp, type = "worst-case")

plot(sense_cust_grp, type = "worst-case",showvars='masked',lim=0.5)

plot(sense_cust_grp, type = "worst-case",showvars='all',lim=0.5)

plot(sense_cust_grp, type = "worst-case",lim=0.5,
     showvars=list('village','villageMngao','age'))

```


